# Introduction {#intro}

Planned chapters

1. Aim of this book
1. Step through a simple example of process of mapmaking [cake first, show achievable]
1. R recap - help ensure readers know foundations, data structures and manipulation
1. Spatial data focusing on real examples : cities, roads, admin boundaries, surfaces  (points lines polygons grids, sf & raster)
1. Visualising spatial data part 1 : sf, mapview, tmap
1. Shiny interactive map applications
1. Sources of data, what is open data (could combine with chap 4)
1. Join non-spatial data onto spatial
1. Raster manipulation, e.g. summarising data by areas
1. OpenStreetMap data, what they are and how to use
1. Creating and editing spatial data
1. Visualising spatial data part 2, more advanced techniques
1. Case study 1 : health facility locations
1. Case study 2 ...


You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
