# A taster - making your first map  {#taster} 

This chapter shows the steps through an example process of map-making. It demonstrates the focus and potential of this book, so that the readers can quickly see what this book aims to achieve and how it is done. 


```{r, c2-packages, results='hide', warning=FALSE, message=FALSE, include = FALSE, eval=FALSE}
#Install packages
cran_packages_to_install <- c(
  "ggplot" ,           # plot-making package
  "tmap",              # map-making package for static and interactive maps
  "tidyverse",         # metapackage containing dplyr, ggplot2 and other packages
  "sf",                # package for working with spatial data
  "s2",                # package for working with spatial data
  "tmaptools",         # map building package
  "raster"
  )

install.packages(cran_packages_to_install)
remotes::install_github("afrimapr/afrilearndata")
```

```{r, c2-libraries, results='hide', warning=FALSE, message=FALSE, include = FALSE}
#Load packages
library(tmap)
library(ggplot2)
library(sf)
library(s2)
library(afrilearndata)
library(dplyr)
library(tmaptools)
library(raster) 
```

```{r c2-dataset, include=FALSE}
#static mode activated
tmap_mode("plot")

```

The `tmap` package is frequently used in geospatial visualisations because it allows us to quickly and easily create both, static and interactive maps. The syntax of `tmap` resembles that of `ggplot2` which is a popular data-visualising package. Both use layers to include additional elements in the map and both follow a strict separation between the data and the aesthetic of the map. Firstly, the dataset is provided in `tm_shape` function and then additional layers of the map are specified through the use of a `+` sign followed by different functions (depending on what we want to add). Below we present two examples of how `tmap` is used to create simple static and interactive maps. In the code below, *africountries* is our dataset, an object of `sf` class, whereas `tm_polygons` and `tm_lines` are used to plot polygons and lines respectively.

## Static maps
```{r c2-simple-map, message=FALSE, warning=FALSE}

sf_use_s2(FALSE) #switches off spherical geometry, needed after sf package update


map_1 = tm_shape(africountries) +
    tm_polygons()

map_2 = tm_shape(africountries) +
    tm_polygons(col = "pop_est", title = "Population")

map_3 = map_1 +
  tm_shape(afrihighway) +
    tm_lines(col = "blue")

tmap_arrange(map_1, map_2, map_3)

```

Furthermore, the creation of an interactive map is equally convenient, where it only requires "turning on" of an interactive mode by changing from "plot" to "view" in the `tmap_mode()` function, as seen in the example below.

## Interactive maps
```{r c2-interactive_map, fig.show="hold", out.width="50%"}
tmap_mode("view")

breaks=c(0,2,20,200,2000,25000)

tm_shape(afripop2020) +
    tm_raster(breaks=breaks, title = "Population per km sq") +
tm_shape(afrihighway) +
    tm_lines(col = "blue") +
tm_shape(africapitals) +
    tm_dots(col = "black", alpha=0.5)
```