# R recap {#rrecap} 
This section provides a brief recap of the basic steps that are required or useful to do before staring on the project.

## Setting working directory
At the beginning of each project it is crucial to determine the working directory for your project. A working directory is a folder where all your files associated with the project will be stored. For example, original datasets, saved datasets as well as plots, graphs or maps created and exported from R. Hence, working directory not only allows you to gather and access the files produced along the work but also load the existing datasets into the RStudio.<br/><br/>

You can set up the working directory with the `openProject` function as follows (recommended):

```{r, eval=FALSE}
rstudioapi::openProject("path to your directory")
```

Alternatively, you can use the following command:
```{r, eval=FALSE}
setwd("path to your directory")
```

While the former option will set up and open the working directory for your project, the latter simply determines the default folder for each specific R script. 

## Creating, naming and saving a new R script

The R script is a plain text file that allows you to save the R code containing both, commands and comments. Saving the R script allows you to reuse your R code and creates a reproducible record of your work. Therefore, it is a good practice to create, name and save it at the very beginning of your project. <br/><br/>
You can crate a new R script by clicking the **New File** icon in the top right corner of the RStudio toolbar, which will open a list with different file options. Choose the **R Script** from the menu and the blank script will open in the main RStudio window.

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: New R script can also be opened by *Shift+Ctrl+N*.
</div>

You can now save your R script by clicking on the Save icon at the top of the **Script Editor**, this will prompt a *Save File* window where you can change the name of your script. Note that the file extension for R scripts is **.R**. At this point you can also choose a folder where to save your new file. By default, RStudio will try to save your new script in the current working directory. Once the name and file location are chosen, simply click the **Save** button.

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: As you work along your R script document, you can quickly save changes by pressing *Ctrl+S*.
</div>

## Executing the code
Now, as you set up the working directory for your project and created your first R script file, we will look into ways of executing (also known as running) the code. You can execute selected chunk of code by clicking **Run** button at the top right corner of the **Script Editor**. 

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: You can also execute the selected code with  *Ctrl+Enter*.
</div>

If you do not select any code and press **Run**, RStudio will run the line of code where the cursor is located. The code which has been exectued will appear in the **Console**, usually located at the bottom part of the window.

In RStudio it is also possible to add comments next to your commands simply using a hashtag (#) beforehand. This will stop R from executing this specific part:
```{r}
3+6 # Using R as a calculator

# 3+6
```

## Install and load packages

## Load data
There are different approaches that can be used to import your datasets, depending on where the data is located. Here we consider two possibilities:

1. Dataset is included in the R package.<br/>
If the dataset is already part of the R package, it will be automatically read in when the package is installed and loaded into R. 
For example:

```{r, eval=FALSE}
install.packages(afrilearndata) #data package
library(afrilearndata)
```

2. Dataset is stored as a separate .txt or .csv file.


## Formats of spatial data (content, extensions) 
## Objects & dataframes
## Saving dataset
## Saving & exporting maps to the file


