# R recap {#rrecap}

Previous section demonstrated what can be achieved with the learning materials provided in this book and gave an example of capabilities of packages that are used. This section presents basic steps that are required (or simply a good practice) before starting the project. It provides a brief recap of R fundamentals such as data types and structures, objects and commands. It aims at building a starting point for those who are new to R and provide a basic background knowledge that the reader can build on, while progressing through the further chapters of this book. It will be particularly useful for beginners, who have minimal knowledge of R. Therefore, more advanced readers can move forward to the next chapter where spatial data types are looked into. <br/>

<style>
div.green { background-color:#e3ffd9; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">
### Learning objectives {-}

- create a new project and start working with data
- identify and characterise different data types and structures in R
- access and search for help within R 

</div>

### List of episodes {-}

1. [Working directory](#wd)
2. [New script](#new_script)
3. [Executing code](#executing)
4. [Install packages](#install_packages)
5. [Load data](#load_data)
6. [Basic data types](#data_basics)
7. [Basic objects](#objects_basics)
8. [Getting help](#help)

## Setting working directory {#wd}
At the beginning of each project it is crucial to determine the working directory for your project. A working directory is a folder where all your files associated with the project will be stored. For example, original data sets, saved data sets as well as plots, graphs or maps created and exported from R. Hence, working directory not only allows you to gather and access the files produced along the work but also load the existing data sets into the RStudio.<br/><br/>

You can set up the working directory with the `openProject` function as follows (recommended):

```{r, eval=FALSE}
rstudioapi::openProject("path to your directory")
```

Alternatively, you can use the following command:
```{r, eval=FALSE}
setwd("path to your directory")
```

While the former option will set up and open the working directory for your project, the latter simply determines the default folder for each specific R script. 

## Creating, naming and saving a new R script {#new_script}

The R script is a plain text file that allows you to save the R code containing both, commands and comments. Saving the R script allows you to reuse your R code and creates a reproducible record of your work. Therefore, it is a good practice to create, name and save it at the very beginning of your project. <br/><br/>
You can crate a new R script by clicking the **New File** icon in the top right corner of the RStudio toolbar, which will open a list with different file options. Choose the **R Script** from the menu and the blank script will open in the main RStudio window.

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: New R script can also be opened by *Shift+Ctrl+N*.
</div>

You can now save your R script by clicking on the Save icon at the top of the **Script Editor**, this will prompt a *Save File* window where you can change the name of your script. Note that the file extension for R scripts is **.R**. At this point you can also choose a folder where to save your new file. By default, RStudio will try to save your new script in the current working directory. Once the name and file location are chosen, simply click the **Save** button.

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: As you work along your R script document, you can quickly save changes by pressing *Ctrl+S*.
</div>

## Executing the code {#executing}
Now, as you set up the working directory for your project and created your first R script file, we will look into ways of executing (also known as running) the code. You can execute selected chunk of code by clicking **Run** button at the top right corner of the **Script Editor**. 

<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: You can also execute the selected code with  *Ctrl+Enter*.
</div>

If you do not select any code and press **Run**, RStudio will run the line of code where the cursor is located. The code which has been exectued will appear in the **Console**, usually located at the bottom part of the window.

In RStudio it is also possible to add comments next to your commands simply using a hashtag (#) beforehand. This will stop R from executing this specific part:

```{r}
3+6 # Using R as a calculator

# 3+6
```

## Install and load packages {#install_packages}
Before being able to use a specific package in R, it is necessary to firstly install and then load it. In this example we install package *remotes* using `install.packages()`:

```{r, eval=FALSE}
install.packages("remotes")
```

Further, we have to load it into R using `library()` function. Note the presence of the quotation mark ("") in case of the first function and its lack in case of the second function.

```{r, eval=FALSE}
library(remotes)
```

To download and install packages directly from remote repositories such as GitHub, we can use previously loaded `remote` package and load them as usual:

```{r, eval=FALSE}
remotes::install_github("afrimapr/afrilearndata")
library(afrilearndata)
```

In the example, above we use double-colon operator *::* to call function `install_github` from name space`remotes`. Only functions which are included in the packages can be retrieved in this manner. 

## Load data {#load_data}
There are different approaches that can be used to import your data sets, depending on where the data is located and in what file type it is stored. The easiest case is when the data set is already part of the R package, as it will be automatically read in when the package is installed and loaded into R.<br/>

For example, install and load packages containing data set:
```{r, eval=FALSE, results='hide', warning=FALSE, message=FALSE}
remotes::install_github("afrimapr/afrilearndata")
library(afrilearndata)
```

To view what data sets are available in a given package:
```{r, c3-read-in-data}
data(package="afrilearndata")
```
We can view and explore the `africapitals` data set using:
```{r, warning=FALSE, message=FALSE}
head(africapitals)
```
We can also create a dataframe from `africapitals` in the global environment:
```{r, warning=FALSE, message=FALSE}
dataset <- africapitals
```

## Basic data types {#data_basics}
In this section we explore a basic set of possible object types that you can encounter in a dataframe. The division presented below is based on the types of values that data (object) stores. The most popular data types are:

- Numeric
- Character
- Logical (so-called Boolean)
- Factor

In R the type of object is referred to as `class` of an object and this function can be used to learn what data types the object contains. 
```{r, warning=FALSE, message=FALSE}
x <- 11 #we create a vector that stores value 11
class(x)
```

It is useful because each class has different properties and can be used in a different way. For example, intuitively we can perform the mathematical operations on numeric objects such that:

```{r, warning=FALSE, message=FALSE}
x <- 11 
y <- 52
x*y # multiplication
```

Hint: To assign a value to an object two operators can be used interchangeably: `<-` or `=`.


### Numeric and integer
*Numeric data* stores real numbers. This means that the object `x` from above is in fact 11.000000, where the zeros are not printed, by default. It is also possible to store value as whole number in a class object called `integer`. An integer can be created from a `numeric` object in the following way:

```{r, warning=FALSE, message=FALSE}
x <- 13.5
z <- as.integer(x)
class(z)
```

Please note, even though the numeric value is a decimal number the integer by default rounds downwards, hence both *13.1* and *13.9* will result in integer equal *13*.

### Characters
Character objects store text, usually referred to as a `string`. String can be a letter, word or even a sentence. 

```{r, warning=FALSE, message=FALSE}
x <- "adult" 
y <- "A"
y <- "I have a bicycle."

```

Interestingly, a character can also contain a number, however it will be stored as a text and will not have the same properties as a `numeric` or `integer` object. As a  results, it will not be possible to perform calculation on character objects, even if they contain numbers. This is when `class` function becomes helpful.

```{r, warning=FALSE, message=FALSE}
x <- "5" 
y <- 7
class(x)
class(y)
```

It is also possible to convert character variables into numeric:

```{r, warning=FALSE, message=FALSE}
x <- "5" 
z <- as.numeric(x)
class(z)
```

### Logical
Logical objects can take values `TRUE` or `FALSE`, where `TRUE` is an equivalent of 1 whereas `FALSE` is an equivalent of 0. In these sense, they can be thought of as numerical values.
```{r, warning=FALSE, message=FALSE}
x <- TRUE
y <- 3 + TRUE
print(y)
```

Typically, logical objects are results of a condition. For example, if we want to test if object `a` is larger than 100.

```{r, warning=FALSE, message=FALSE}
a <- 76 #create object a
a > 100 # condition 1
a < 80 # condition 2
```

### Factors
Factors are categorical variables with associated `levels`. They can store both, numbers:
```{r, warning=FALSE, message=FALSE}
a <- rep(1:3, times=3) # create a vector of numbers from 1 to 3, repeated 3 times
a <- as.factor(a)
a
```
and strings:
```{r, warning=FALSE, message=FALSE}
b <- c("A", "B", "B", "C", "D", "C", "C")
b <- as.factor(b)
b
```

The numbers and strings, in the example above, serve as labels of different levels.

### Exercises
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 3.6.1**: Create a vector 'a' that contains descending sequence of numbers from 50 to 30 that decrases by 2. Hint: use seq()
</div>  

## Basic objects and data structures {#objects_basics}
In this section we explain the basic data structures that we often work with in R. These include a vector, list, matrix and a data frame.

### Vector
Vector is a one dimensional structure which contains elements of the same type. Usually a `combine` function is used to create a vector such that:
```{r, warning=FALSE, message=FALSE}
a <- c(1:10)
```

It is also possible to create a vector with text-based objects, for example with capitals from our `afrilearndata`:
```{r, warning=FALSE, message=FALSE}
capitals <- africapitals$capitalname
```

We can `print` content of our vector to see what it contains:
```{r, warning=FALSE, message=FALSE}
print(capitals)
```

Another important feature is to see what types of objects the vector stores. We can see that our text-based objects are *characters* using `typeof` function:
```{r, warning=FALSE, message=FALSE}
typeof(capitals)
```

### Matrix
Matrix can be though of as an upgraded version of a vector, where vector is an one-dimensional array and two-dimensional array. We can create a matrix that has three columns and five rows with the following:
```{r, warning=FALSE, message=FALSE}
matrix(1:15, ncol=3, nrow=5, byrow = TRUE)
```

By default, the elements in the matrix will be arranged by column, so `byrow=TRUE` segregates the elements in the matrix row-wise.<br/>

We can also exploit the fact that matrix is an upgraded vector by binding them column- or row-wise, as long as they have the same length:
```{r, warning=FALSE, message=FALSE}
vector1 <- c("water", "milk", "juice")
vector2 <- (1:3)
cbind(vector1, vector2) #for row-wise binding we would use rbind()
```

### Data frame
Data frame is a two dimensional array, similar to a matrix. However, it can contain different data types, unlike matrix. Beyond, columns as well as rows within the data frame are named. In a data frame a column is a variable and row is an observation. You can create a simple data frame with two columns and three rows, using vectors above, with `data.frame` function:
```{r, warning=FALSE, message=FALSE}
data.frame(vector1, vector2)
```

If you compare the output of the example above where we used the `cbind` function to create a matrix, to the result of the `data.frame` function, we can clearly see the difference between a matrix and a data frame. Where matrix is homogeneous and data frame is heterogeneous in terms of data type. All the values in the matrix are characters while the first column in data frame contains factors and the second one integers. <br/>

To address a specific column in a data frame we can use *$* such that:
```{r, warning=FALSE, message=FALSE}
df_1 = data.frame(vector1, vector2) #create data frame called df_1
df_1$vector1 #view column called vector1
```

At the moment columns in the *df_1* adapted names of the vectors:
```{r, warning=FALSE, message=FALSE}
colnames(df_1)
```

We can change the column headers by their *names* or their *index* such that:
```{r, warning=FALSE, message=FALSE}
names(df_1)[names(df_1) == "vector1"] <- "drinks"
names(df_1)[2] <- "amount"
colnames(df_1)

```

### Exercises
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 3.7.1**: Create three vectors containing character, integer and logical values, respectively. Each with 10 rows. Bind them to generate a matrix with 3 columns and 10 rows.
</div>  
<br/>
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 3.7.2**: Create a data frame from the matrix created in the previous exercise. Rename column headers to describe their content.
</div>  
<br/>

## Getting help with R Help {#help}
R is a powerful software with many different packages and functions which are continuously developed and added. Therefore, it is difficult, if not impossible, to be familiar with all the functions that are currently available. Hence, in order to learn more about them, R provides extensive documentation which can be accessed with `help()` function, `? operator` or by clicking in *Help* panel in the bottom-right corner. For example:
```{r, warning=FALSE, message=FALSE}
help(matrix) #to find out what matrix is

help(package="remotes") #to get help with remotes package
```

Alternatively, it is possible to use `?` in front of the searched item and the R documentation will appear in right-bottom pane. For example, to read more on how to use help function itself we can use: 
```{r, warning=FALSE, message=FALSE}
?help 
```

Importantly, in order to get help regarding objects with unusual names such as the logical operators, it is necessary to use quotation marks:
```{r, warning=FALSE, message=FALSE}
?"&"

help("!")
```

R documentation for functions frequently offers a working example which can be accessed with:
```{r, warning=FALSE, message=FALSE}
example(vector) #to see an example of a vector
```

## Further resources
Finally, the Internet also provides several reliable sources such as official [R website](r-project.org) or [documentation pages](rdocumentation.org) where you can find help. 

## Wrap up
In this chapter we learned several R commands which prepared the reader for starting with his/her own project. Moreover, we familiarised with a number of data types and R objects. Moreover, we looked at how to obtain help and access R documentation. The material covered in this section aimed at building a base for the reader to allow him/her successfully progress through the book. 

## Exercise solutions {#solutions_ch3}

- Exercise 3.6.1
```{r}
a <- seq(50,30, by = -2)
```

- Exercise 3.7.1
```{r, warning=FALSE, message=FALSE}
v1 <- c("water", "milk", "juice", "coffee", "tea", "tea", "juice", "milk", "water", "soda")
v2 <- seq(1,10)
v3 <- rep(c(TRUE,FALSE), length.out = 10) 
v3 <-as.logical(v3)
m1 = cbind(v1, v2, v3) 
m1
```
- Exercise 3.7.2
```{r, warning=FALSE, message=FALSE}
df = data.frame(m1)

names(df)[1] <- "Drink"
names(df)[2] <- "Amount"
names(df)[3] <- "Female"

df$Female <-as.logical(df$Female)
df$Amount <-as.integer(df$Amount)
df
```
