# Getting data into R {#getting}

## Overall goal of the chapter


<style>
div.green { background-color:#e3ffd9; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">
**Learning objectives**
- Familiarise with different data sources and methods of reading in datasets
- Be able to read in data of the 4 types into R from files and other sources

</div>


This chapter requires several packages, some of them we mioght have used before. Let's install and load them.

```{r, eval=FALSE}
install.packages("sf")             
install.packages("afrilearndata")    
install.packages("raster")           
install.packages("mapview")          
install.packages("readr")            
install.packages("rgdal") 
```

```{r, results='hide', warning=FALSE, message=FALSE}
# library(sf)               # working with vector data
# library(afrilearndata)    # example data
# library(raster)           # raster manipulation
# library(mapview)          # interactive mapping
# library(readr)            # reading text files
# library(rgdal)            # to try to correct raster error on shinyapps
```

