# Visualising spatial data part 1: tmap and mapview {#visualising1}

## Overall goal of the chapter
In the previous section we explored different types of spatial data and looked at how these can be used to create a map. In this section we will use `tmap` and `mapview` packages to crate both static and interactive maps. In particular, we will focus on the goals presented below.

<style>
div.green { background-color:#e3ffd9; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">
**Learning objectives**

- Create static and interactive maps using different packages
- Learn function structure to build more advanced maps
- Analyse the differences between used packages and identify their applicability for various purposes

</div>


This chapter requires the following packages (some of them we used already):

```{r, eval=FALSE}
cran_packages_to_install <- c(
  "mapview",           # map-making package for interactive maps
  "tmap",              # map-making package for static and interactive maps
  "tidyverse",         # metapackage containing dplyr, ggplot2 and other packages
  "sf",                # package for working with spatial data
  "tmaptools"          # map building package
  )
install.packages(cran_packages_to_install)
#remotes::install_github("afrimapr/afrilearnr")
remotes::install_github("r-spatial/mapview@develop")
remotes::install_github("afrimapr/afrilearndata")
```

Next, we need to load the packages:

```{r, results='hide', warning=FALSE, message=FALSE}
library(tmap)
library(mapview)
library(ggplot2)
library(sf)
library(afrilearndata)
library(dplyr)
library(tmaptools)
#library(afrilearnr)
```

## Quick interactive maps with mapview
`Mapview package` allows quickly and easily create interactive maps of spatial data. Therefore, it is useful when we want to quickly explore the data visually without excessive care about the quality of the presentation. However, if we need static maps or interactive maps which are more elaborate we might want to consider other options such as `tmap package` which will be presented later in this chapter. <br/> 
However, a potentially important aspect to consider is whether the map should be interactive at all. Even though powerful R packages allow us to produce interactive maps quickly and (nearly) effortlessly, it is important to align the purpose of the map with its appearance and functionality. This will ensure that the interactivity does not impede reader's interaction with the map and helps conveying key information easily.For example, it is important to consider target audience of the map. If the map is for the specialists then the interactivity will potentially help in more in-depth analysis of the data and finding more detailed, micro-scale information, beyond the trends identifiable at first sight. However, if the map is published for broader audience to provide an overview of a certain situation easily and quickly, then a static map is probably a better medium. At the end of the day the features included in the map should help the author of the map communicate with the reader. Therefore, the type, functions and appearance of the map should be in line with the amount and level of information to be presented, the purpose of the map and level of expertise of the reader.<br/>

As we now shortly discussed the advantages and disadvantages of using the interactive maps, let us have a look on the how they can be created. Firstly, it is important to say that the typical spatial objects types supported by `mapview` package are: 

- sf
- raster
- sp

To create the most basic map with the boundaries of African countries we call `mapview()` function

```{r c5-mapview-countries, warning=FALSE, message=FALSE}
 data(africountries)
 mapview(africountries)
```

This map has a number of functions:

- layer button to switch between five different layers
- scale bar
- zoom butttons
- data name in the top right corner

We can easily fill an interactive map with the colours based on one of the columns. Here we use population column to colours our map. Importantly, when we hoover over specific country the population data for a given country is showed.


```{r c5-mapview-countries2}
mapview(africountries, zcol = ("pop_est"))
```

Similarly, we can control for other features such as line width (lwd), colour of polygons (col.regions). Another important feature that we might want to control is the underlying map which we set with 'map.types' argument. The available background maps can be viewed [here](http://leaflet-extras.github.io/leaflet-providers/preview/).


```{r c5-mapview-countries3}
mapview(africountries, color = "red", 
      lwd = 1.5, col.regions = "white", map.types = "Esri.WorldShadedRelief")
```

Further, mapview allows us to easily build up many layers within a single map. Here, we firstly include countries and then add location of airports. Further a useful option is `layer.name` which allows us to control the naming of different layers of the map. The specific layers can be controlled and adjusted as showed in the example. In partucular, we set the colour of countries polygons according to the population size. It is possible to use layer button to hide and unhide layers of the map.

```{r c5-mapview-airports}
# simple map with two layers
mapview(list(africountries,afriairports), layer.name = c("Countries", "Airports"))

#map with advanced layers options
mapview(africountries, zcol = "pop_est", layer.name = "Population") +
  mapview(afriairports, layer.name = "Airports location")
```

The link to full documentation is available in *Further resources* section. 

### Exercises
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 1**: Create a map with two layers: countries and capitals. Fill the country polygons according to income group. Explore the layers button to hide and unhide some of them.  
</div>   

## Static maps with tmap
In this section we explore the potential of `tmap` in creating static maps, which is a flexible, yet user-friendly package. Its syntax is similar to `ggplot2` (plot-making package), where additional features of the map are simply build up on top of the basic structure. Before we plot the first maps we need to create a dataset called *africa* and activate static map viewing mode.

```{r c5-tmap-countries1, message=FALSE}
tmap_mode("plot") #activates static map viewing mode
```

Next, we create an empty contour map of the continent, where `tm_shape()` function serves as a basic component containing the shape object i.e. our spatial dataset. On top of this basic element we can then add more layers using `+` operator. The advantage of `tmap` package is that it offers a large variety layers and these are named relatively intuitively to make the search and use easier. For example, we will now use layer called `tm_borders()` which defines the borders of the polygons. Further, we create an empty, borderless map using `tm_fill()` which defines the fill of the polygons. 

```{r c5-tmap-countries2, fig.show='hold', out.width="50%"}

tm_shape(africountries)+
 tm_borders()

tm_shape(africountries)+
 tm_fill()

```

Intuitively to create a map with both layers, we need to include `tm_fill()` and `tm_borders()` simultaneously. Alternatively, `tm_polygons()` function allows us to achieve exactly the same result more efficiently. It draws the polygons borders and fills them.

```{r c5-tmap-countries4,fig.show="hold", out.width="50%"}

tm_shape(africountries)+
 tm_borders()+
  tm_fill()

tm_shape(africountries)+
 tm_polygons()

```

We can further develop our map by adjusting different features of the map such as colour and transparency of the map, width and types of the border lines etc. The extra parameters for changing the plot elements can be found [here](http://publish.illinois.edu/johnrgallagher/files/2015/10/BaseGraphicsCheatsheet.pdf).


```{r c5-tmap-countries5, out.width="50%"}

tm_shape(africountries)+
 tm_polygons(col = "green", lty = "dotted", alpha = 0.3)

```

As we already know a basic structure required to build a static map, we can proceed to a more advanced (and useful) example, where we fill the polygons using data on **last census** and adding relevant title to the legend. Further, we can build on the previous map by adding labels for country names with `tm_text` function. In a similar manner a variety of different layers can be added. 

```{r c5-tmap-countries6, fig.show='hold', out.width="50%", fig.cap="Basic map without labels (left) and map with added coutries names (right)"}

tm_shape(africountries)+
 tm_polygons(col = "lastcensus", title = "Year of last census")+
  tm_layout(scale = 1.25) # increases the overall size of the legend

tm_shape(africountries)+
 tm_polygons(col = "lastcensus", title = "Year of last census")+
 tm_text("name", size = 0.52, fontface = "bold", auto.placement = TRUE)+
  tm_layout(scale = 1.25) 

```

Another useful function in `tmap` is the adjustment of the intervals in the legend, in case the default scale does not provide enough variety. For example, plotting the population of African countries on the map using default intervals (stemming from the data) is not very informative due to to much clustering of information. Instead, we might want to use `breaks` argument to set them up manually to provide more differentiation visually. For clarity, I set up the customised breaks outside of the map-making function. Alternatively, we can control number of so-called `bins` (groups) into which the **Population** is divided, by setting `n` argument to a desired number of bins inside the `tm_polygons` function.

```{r c5-tmap-countries-breaks,warnings = FALSE, message = FALSE,out.width="75%"}
#default settings
tm_shape(africountries) +
                tm_polygons(col = "pop_est", title = "Population")

#set up breaks manually
custom_breaks = c(0, 1, 4, 7, 10, 30, 50, 70, 100, 120, 150, 180, 200) * 1000000
tm_shape(africountries) +
              tm_polygons(col = "pop_est", title = "Population", breaks = custom_breaks)

#set up a number of bins
tm_shape(africountries) +
            tm_polygons(col = "pop_est", title = "Population", n = 15) +
            tm_layout(scale = 0.9)


```

So far, we used continuous variables to fill the polygons but it is also possible to use categorical variables for this purpose. In the example below *income_grp* variable is a categorical variable with income level. It allows for visual grouping of countries in the same category. The names of the categories in the legend were adjusted using `labels` argument. In the similar manner we can show the airports of Africa according to their size.


```{r c5-tmap-countries-categorical, out.width="50%"}
tm_shape(africountries)+
  tm_polygons(col = "income_grp", title = "Income levels", alpha = 0.6,
              labels = c("High", "Upper middle", "Lower middle", "Lower"))+
                tm_layout(scale = 1.25)

data(afriairports)
tm_shape(africontinent)+
  tm_polygons(col = "lightyellow")+
tm_shape(afriairports) +
  tm_dots(col = "type", shape = 21, size = 0.2, title = "Airport size",
          labels = c("Large", "Medium", "Small"),
          palette=c(large_airport='red', medium_airport='yellow', small_airport='blue')) +
            tm_layout(scale = 1.25)

```

It is important to bear in mind that same information can be displayed on the map in many different ways. For example, we can make use of 'tm_bubbles' to create so called bubbles which size is linked to the population of each country while their colour related to the income group.


```{r c5-tmap-bubbles, out.width="75%"}
tm_shape(africountries)+
 tm_polygons(col = "lightyellow") +
  tm_bubbles(size = "pop_est", col = "income_grp", 
                    title.col = "Income level", title.size = "Population")

```

### Colour setting and colour pallete

An important aspect of the maps, as in every visual representation, are the colours. In `tmap` we can change the default colours using `palette` argument. In the example below, we used palette called **Set1** and **plasma** and there are many other options to choose from. To view them, we run `tmaptools::palette_explorer()` which will open a new window with a wide range of palettes. Sliders on the left hand side allow selecting a number of colours.


```{r c5-tmap-palette, warning=FALSE, message = FALSE, results='hide', fig.cap = "Africa's maps using different pallete options."}

set1 = tm_shape(africountries)+
         tm_polygons(col = "lastcensus", title = "Year of last census", palette = "Set1")

plasma = tm_shape(africountries)+
          tm_polygons(col = "lastcensus", title = "Year of last census", palette = "plasma")

tmap_arrange(set1, plasma)

```

Alternatively an overall style of a single map can be easily changed using `tm_style` function. However, it might also be useful to change the style in all the maps in the document. This can be achieved with setting the style globally with `tmap_style("nameofthestyle")`. The default style is called "white", hence to restore a default style setting use `tmap_style("white")`. 

```{r, include = FALSE}
tmap_mode("plot")
```


```{r c5-tmap-style,message = FALSE, fig.cap = "Examples of different styles."}

classic = tm_style("classic")+
            tm_shape(africountries)+
              tm_polygons()+
                tm_bubbles(size = "pop_est", col = "income_grp", title.col = "Income level",
                  title.size = "Population", labels = c("High", "Upper middle", "Lower middle", "Lower"))+
                    tm_layout(scale = 0.8)

beaver = tm_style("beaver")+
          tm_shape(africountries)+
            tm_polygons()+
              tm_bubbles(size = "pop_est", col = "income_grp", title.col = "Income level",
                title.size = "Population", labels = c("High", "Upper middle", "Lower middle", "Lower"))+
                  tm_layout(scale = 0.8)

albatross = tm_style("albatross")+
            tm_shape(africountries)+
              tm_polygons()+
                 tm_bubbles(size = "pop_est", col = "income_grp", title.col = "Income level",
                  title.size = "Population", labels = c("High", "Upper middle", "Lower middle", "Lower"))+
                    tm_layout(scale = 0.8)

tmap_arrange(classic, beaver, albatross)

```

### Visualising different datasets in one map
`tmap` not only allows us to add extra layers to a single map by building on a basic structure, but it also enables us to join two separate datasets in a single map. We first create an object called *countries* which is a map with African countries:
```{r c5-tmap-combine1, warning=FALSE, out.width="75%"}
#read in data
data(africountries)
#create an object
countries = tm_shape(africountries)+
  tm_polygons(col = "lightblue", alpha = 0.3)
#view object called "countries"
countries
```

Then we use it as an argument to merge it with a map containing the location of capitals.

```{r c5-tmap-combine2, warning=FALSE, out.width="75%"}
#read in data with capitals
data(africapitals)
#merge two maps
countries+
  tm_shape(africapitals)+  #uses dataset with capitals
    tm_dots(col = "red", size = 0.15)+       #adds dots on the map
      tm_layout(title = "Capital cities of Africa" , title.position = c("right", "top"), title.size = 1) #add and adjust map title

```

### Exercises
The solution to the exercises are provided at the [end of the chapter](#solutions_ch4).

<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 2**: Plot a yellow map of Africa with transparency of 0.4 and solid line borders of width 3. 
</div>   
<br/>
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 3**: Plot a lightblue map of Africa with dotted, non-transparent, red borders.
</div>  
<br/>
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 4**: Plot a map that has polygons filled with gross domestic product, black borders, legend titled "GDP" and customised breaks at 0, 20000, 30000, 40000, 50000, 100000, 200000, 300000, 500000.
</div>  
<br/>
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 5**: Create a map with continent contour in light green colour. Then add african highways by using *afrihighway* dataset. Finally, include a title on the bottom of the map. 
</div>  

## Interactive maps with tmap
The `mapview` is not package that allows us to produce interactive maps. Another possibility is the `tmap`package that offers versatility not only for creating static maps, as we have seen above, but it also allows for making customised interactive maps conveniently by simply changing the mode of operation from static to dynamic using *plot* or *view*, respectively in `tmap_mode()` function. After the activation of an interactive mode all the maps produced with `tmap` will be interactive. Therefore, we are able to create all the maps that we have produced so far in an interactive version.
```{r c5-tmap-interactive, warning=FALSE}
#sets interactive mode
tmap_mode("view")

tm_shape(africountries)+
  tm_borders()+
    tm_shape(africapitals)+
    tm_dots(col = "white", border.col = "blue", size = 0.1)+ #capital dots settings
    tm_text("capitalname")+ #adds capital names
        tm_shape(afrihighway)+
    tm_lines(col = "red", lwd = 2, alpha = 0.7) #highway lines setting

```

Regardless of the mode, the creation of maps using `tmap` package is easy and convenient. The additional feature in the interactive mode is the choice of the background map. This can be set up using function `tm_basemap`. Available base maps can be accessed using `leaflet::providers`. They can also be previewed [here](https://leaflet-extras.github.io/leaflet-providers/preview/). In the example below, we use a topographic map of Africa. 


```{r c5-tmap-basemap}
tm_basemap("OpenTopoMap")+
tm_shape(africountries)+
  tm_borders()+
    tm_shape(africapitals)+
    tm_dots(col = "black")+ #capital dots settings
        tm_shape(afrihighway)+
    tm_lines(col = "red", lwd = 2, alpha = 0.7) #highway lines setting

```

### Visualising subsets of data

Let's assume that we are only interested in the subset of our data for example a specific country or the location of large airports only. We can achieve it using `dplyr` package. 

```{r c5-tmap-subset}
tmap_mode("view")
data(afriairports)

#take a subset of data
large = afriairports %>% filter(type=="large_airport")

#plot the subset
tm_shape(africountries)+
  tm_borders()+
  tm_shape(large)+
    tm_dots( col = "red")

```

### Exercises

<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 6**: Create an interactive map uses watercolour theme, the highways are blue and dots denoting capitals are white with black borders. Finally, add capitalnames.
</div>  
<br/>
<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 7**: Create an interactive map of African countries which have population smaller than 10 million. Fill the polygons with the population size.
</div>  

## Further resources
- [tmap documentation](https://cran.r-project.org/web/packages/tmap/tmap.pdf)
- [Mapview documentation](https://r-spatial.github.io/mapview/articles/)

## Summary
In  this chapter we looked at how to produce static and interactive maps using different packages. Specifically, we saw that the `tmap` package is a flexible tool to create maps with a stepwise approach, where we can add different layers to our maps. It is equally easy to switch between static and interactive maps. Further, `mapview` was showed to be a quick solution for creating interactive maps.


## Exercise solutions {#solutions_ch4}
```{r, include = FALSE}
tmap_mode("plot")
```

- Exercise 1

```{r c5-ex-5.2.1, out.width="60%"}
mapview(africountries, zcol = "income_grp") + mapview(africapitals)
```

- Exercise 2
```{r c5-ex-5.3.1, out.width="60%"}
 tm_shape(africountries)+
  tm_polygons(col = "yellow", lwd = 3, lty = "solid", alpha = 0.4)
```

- Exercise 3
```{r c5-ex-5.3.2, out.width="60%"}
tm_shape(africountries)+
 tm_fill(col = "lightblue")+
 tm_borders(col = "red", lty = "dotted", alpha = 1)
```

- Exercise 4
```{r c5-ex-5.3.3, out.width="60%"}
custom_breaks = c(0, 2, 3, 4, 5, 10,20, 30, 50) * 10000
tm_shape(africountries)+
 tm_polygons(col = "gdp_md_est", title = "GDP",  breaks = custom_breaks)
```

- Exercise 5
```{r c5-ex-5.3.4, out.width="60%"}
map_contour = tm_shape(africountries)+
  tm_fill(col = "lightgreen")
data(afrihighway)
map_contour +
  tm_shape(afrihighway)+
  tm_lines()+
  tm_layout(title = "Highways in Africa", title.position = c("center", "bottom"))

```

- Exercise 6

```{r c5-ex-5.4.1, out.width="60%"}
tmap_mode("view")


tm_basemap("Stamen.Watercolor")+
tm_shape(africountries)+
  tm_borders()+
    tm_shape(africapitals)+
    tm_dots(col = "white", border.col = "black")+ #capital dots settings
    tm_text("capitalname", size = 1.2)+
        tm_shape(afrihighway)+
    tm_lines(col = "blue", lwd = 3) #highway lines setting

```

- Exercise 7

```{r c5-ex-5.4.2}
tmap_mode("view")
data(africountries)

#take a subset of data
small = africountries %>% filter(pop_est < 10000000)

tm_shape(small)+
  tm_polygons(col = "pop_est")

```
