# Visualising spatial data part 1: sf, mapview, tmap {#visualising-1}

In this section we will learn:

1. How to create a static map 
    + Filled with colours based on the categorical variable
    + Filled with colours based on the continuous variable
2. How to create an ordered and unordered list

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
**Hint**: In order to work properly a list requires an empty line beforehand.
</div>

## tmap package
### Basics (this step can be removed if the basic steps can be put in ch.3 'R recap', since they repeat for all packages)
Set up the working directory 
```{}
setwd("path to your directory")
```

Install required packages

```{r, results=FALSE, message=FALSE, warning=FALSE}
install.packages(c("ggplot2","sf", "rnaturalearth", "rnaturalearthdata"))
```

Load the packages

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
```

### Static maps

### Layers
### Interactive maps

## sf package 

## mapview package
Mapview package allows for an interactive viewing of spatial objects in R.

