# Visualising spatial data part 1: tmap and mapview {#visualising1}

In this section we will visualise spatial data, in particular we will use `tmap` and `mapview` to crate both static and interactive maps. This chapter requires the following packages (some of them we used already):

```{r, eval=FALSE}
cran_packages_to_install <- c(
  "mapview"            # map-making package
  "tmap"               # map-making package
  "tidyverse",         # metapackage containing dplyr, ggplot2 and other packages
  "sf",                # package for working with spatial data
  "rnaturalearth",     # data package
  "rnaturalearthdata"  # data package
  )
install.packages(cran_packages_to_install)

remotes::install_github("afrimapr/afrilearndata")
```

Next, we need to load the packages:

```{r, results='hide', warning=FALSE, message=FALSE}
library(tmap)
library(mapview)
library(ggplot2)
library(sf)
library(afrilearndata)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
```



## Quick interactive maps with mapview
Mapview package allows for an interactive viewing of spatial objects in R.

```{r, warning=FALSE, message=FALSE}
#mapview(africa)
```

## Layers for the interactive maps
## Static maps with tmap
Create a dataset
```{r}
africa <- ne_countries(continent = 'africa', scale = "medium", returnclass = "sf")
```

Create an empty contour map of the continent

```{r}
tm_shape(africa)+
  tm_borders()
```

## Layers for the static maps

Fill map using data on last census data and add labels for country names

```{r, warning=FALSE}
tm_shape(africa)+
  tm_fill(col="lastcensus", title="Year of last census", legend.format = list(format="s"))+
  tm_borders()+
  tm_text("admin", size = 0.52,  fontface="bold", auto.placement = TRUE)

```

## Interactive maps with tmap


