# Visualising spatial data part 1: sf, mapview, tmap {#visualising-1}

In this section we will learn:

1. How to create a static map 
    + Filled with colours based on the categorical variable
    + Filled with colours based on the continuous variable
2. How to create an ordered and unordered list

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
**Hint**: In order to work properly a list requires an empty line beforehand.
</div>

## tmap package
### Basics (this step can be removed if the basic steps can be put in ch.3 'R recap', since they repeat for all packages)
Set up the working directory with the command `setwd("path to your directory")` or with the `openProject` function as follows (recommended):

```{r, eval=FALSE}
rstudioapi::openProject("path to your directory")
```

Install required packages

```{r, eval=FALSE}
cran_packages_to_install <- c(
  "tmap"               # map-making package
  "tidyverse",         # metapackage containing dplyr, ggplot2 and other packages
  "sf",                # package for working with spatial data
  "rnaturalearth",     # data package
  "rnaturalearthdata"  # data package
  )
install.packages(cran_packages_to_install)
```

Load the packages

```{r, results='hide', warning=FALSE, message=FALSE}
library(tmap)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
```

Create a dataset

```{r}
africa <- ne_countries(continent = 'africa', scale = "medium", returnclass = "sf")
```

### Static maps

Create an empty contour map of the continent

```{r}
tm_shape(africa)+
  tm_borders()
```

Add fill based on last census data and add labels for country names

```{r}
tm_shape(africa)+
  tm_fill(col="lastcensus", title="Year of last census", legend.format = list(format="s"))+
  tm_borders()+
  tm_text("admin", size = 0.52,  fontface="bold", auto.placement = TRUE)

```

### Layers

### Interactive maps

## sf 

## mapview 

Mapview package allows for an interactive viewing of spatial objects in R.

