# Visualising spatial data part 1: sf, mapview, tmap {#visualising-1}

In this section we will learn:

1. How to create a static map 
    + Filled with colours based on the categorical variable
    + Filled with colours based on the continuous variable
2. How to create an ordered and unordered list

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
**Hint**: In order to work properly a list requires an empty line beforehand.
</div>

## tmap package
### Basics (this step can be removed if the basic steps can be put in ch.3 'R recap', since they repeat for all packages)
Set up the working directory 
```{}
setwd("path to your directory")
```
Install required packages
```{r results='hide'}
install.packages(c("ggplot2","sf", "rnaturalearth", "rnaturalearthdata"))
```
Load the packages
```{r results='hide'}
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library("dplyr")
```
Create a dataset

### Static maps

### Layers
### Interactive maps

## sf package 

## mapview package
Mapview package allows for an interactive viewing of spatial objects in R.

