# Visualising spatial data part 1: sf, mapview, tmap {#visualising-1}

In this section we will learn:

1. How to create a static map 
    + Filling with colours based on the categorical variable
    + Filling with colours based on the continuous variable
2. How to create an ordered and unordered list

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
**Hint**: In order to work properly a list requires an empty line beforehand.
</div>
## tmap package
### Basics (this step can be removed if the basic steps can be put in ch.3 'R recap', since they repeat for all packages)
Set up the working directory 
```{}
setwd("path to your directory")
```
Install required packages
```{r results='hide', message=FALSE, warning=FALSE}
install.packages(c("ggplot2","sf", "rnaturalearth", "rnaturalearthdata"))
```
Load the packages
```{r results='hide', message=FALSE, warning=FALSE}
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library("dplyr")
```
Create a dataset
```{r results='hide', message=FALSE, warning=FALSE}
africa = ne_countries(continent = 'africa', scale = "medium", returnclass = "sf")
```
### Static maps
Below we will see how to fill the map with colours based on the continuous and categorical variables. First, it is useful to check the class of *last census* variable.
```{r}
class(africa$lastcensus)
```
```{r map1-fig, fig.cap='Africa with continuous colour fill', out.width='80%', fig.asp=.75, fig.align='center'}
ggplot(africa) + 
  geom_sf(aes(fill=lastcensus)) + 
  coord_sf(datum = NA) + 
  labs(fill = "Last census")+
  theme_minimal()
```
### Layers
### Interactive maps

## sf package 

## mapview package
Mapview package allows for an interactive viewing of spatial objects in R.

