# Sources of data {#data-sources}

Here we show several R packages that can be used to download spatial data including administrative boundaries, population, climatic, and health data.
We give short reproducible examples on how to download and visualize spatial data that can be useful in different settings.
More extended examples and details about the capabilities of each of the packages can be seen at https://rspatialdata.github.io/ and the packages' websites.


## Administrative boundaries

We can download administrative boundaries of countries with the `rgeoboundaries` package.
For example, we can download the administrative boundaries of Nigeria with `geoboundaries()` as follows.

```{r}
# install.packages("rgeoboundaries")
# library(rgeoboundaries)
# library(ggplot2)
# map <- geoboundaries("Nigeria")
# ggplot(data = map) + geom_sf()
```

We can also download administrative boundaries of multiple countries and different administrative levels.

```{r}
# map <- geoboundaries(c("Nigeria", "Chad"), "adm1")
# ggplot(data = map) + geom_sf()
```

## Population

The `wopr` package provides access to the [WorldPop Open Population Repository](https://wopr.worldpop.org/) and provides estimates of population sizes for specific geographic areas.

To download the population data we first use `getCatalogue()` to retrieve the WOPR data catalogue and see a list of currently available databases.

```{r}
install.packages("remotes")
library(remotes)
remotes::install_github('wpgp/wopr')
library(wopr)
catalogue <- getCatalogue(spatial_query = T)
catalogue
```

Then, we subset the catalogue for the country (e.g. Nigeria with ISO code NGA) and download the data for the selection using the `downloadData()` function.

```{r, eval = FALSE}
selection <- subset(catalogue, country == "NGA" & category == "Population" & version == "v1.2")
downloadData(selection)
```

Finally, we use the `sf` package to read in the downloaded shapefile and plot it with `ggplot()`.

[Paula: I've set this to eval=FALSE for now to avoid filenotfound error, can we make this runnable ?, Andy]
```{r c7-data-NGA}
library(sf)
library(ggplot2)

map <- st_read("data/NGA_population_v1_2_admin_level2_boundaries.shp")

 ggplot(map, aes(fill = mean)) + geom_sf() +
   scale_fill_continuous(name = "Population")
```




## Open Street Map data

We can retrieve [Open Street Map data](https://www.openstreetmap.org/) using the `osmdata` package. The `available_features()` function can be used to get the list of recognized features in OSM.
The `available_tags()` function lists out the tags associated with each feature.
A list of the available features can be found in the [OSM wiki](https://wiki.openstreetmap.org/wiki/Map_Features).

```{r}
library(osmdata)
head(available_features())
head(available_tags("amenity"))
```

The first step in creating an `osmdata` query is defining the geographical area we wish to include in the query. This can be done by defining a bounding box that defines a geographical area by its bounding latitudes and longitudes.

```{r}
# lagos_bb <- getbb("Lagos")
# lagos_bb
```

To retrieve the required features of a place (defined by the bounding box), we have to then create an overpass query with `opq()`.
Then, the `add_osm_feature()` function can be used to add the required features to the query. Finally, we use the `osmdata_sf()` function to obtain a simple feature object of the resultant query.

```{r}
# lagos_hospitals <- lagos_bb %>% opq() %>%
#   add_osm_feature(key = "amenity", value = "hospital") %>%
#   osmdata_sf()
```

We can visualise an interactive map of the hospitals in Lagos as follows.

```{r}
# library(leaflet)
# leaflet() %>%  addTiles() %>%
#   addPolygons(data = lagos_hospitals$osm_polygons,
#               label = lagos_hospitals$osm_polygons$name)
```



## Elevation

The `elevatr` package can be used to download elevation data.
We can use `get_elev_raster()` to obtain the elevation of Ghana as follows.


```{r}
# install.packages("elevatr")
# library(ggplot2)
# library(raster)
# library(elevatr)
# library(rgeoboundaries)
# library(sf)
# library(viridis)
# 
# map <- geoboundaries("Ghana")
# 
# elevation_data <- get_elev_raster(locations = map, z = 9, clip = "locations") 
# 
# elevation_data <- as.data.frame(elevation_data, xy = TRUE)
# colnames(elevation_data)[3] = "elevation"
# #remove rows of data frame with one or more NA's using complete.cases
# elevation_data <- elevation_data[complete.cases(elevation_data), ] 
# 
# ggplot() + geom_raster(data = elevation_data, aes(x = x, y = y, fill = elevation)) +
#   geom_sf(data = map, color = "white", fill = NA) +
#   coord_sf() + scale_fill_viridis_c() + ggtitle("Elevation")

```

      

## Temperature

The `raster` package provides access to the [WorldClim](https://www.worldclim.org/) database, and allows us to download data sets on the many different climatic conditions.

For example, we can can download maximum temperature at a resolution of 10 minutes as follows.

```{r}
# library(raster)
# library(ggplot2)
# 
# tmax_data <- getData(name = "worldclim", var = "tmax", res = 10)
# 
# # Degrees Celsius
# gain(tmax_data) <- 0.1
# 
# # Converting the raster object into a dataframe
# tmax_data_may_df <- as.data.frame(tmax_data$tmax5, xy = TRUE, na.rm = TRUE)
# rownames(tmax_data_may_df) <- c()
# 
# # Plot
# ggplot(data = tmax_data_may_df, aes(x=x,y=y)) + geom_raster(aes(fill = tmax5)) +
#   labs(title = "Maximum temperature in May", subtitle = "For the years 1970-2000") +
#   xlab("Longitude") + ylab("Latitude") +
#   scale_fill_gradientn(name = "Temperature (°C)", colours=c("#0094D1","#68C1E6", "#FEED99","#AF3301"), breaks = c(-20, 0, 20, 40))
```


## Rainfall

The `nasapower` package aims at making it quick and easy to automate downloading [NASA POWER (NASA Prediction of Worldwide Energy Resource)](https://power.larc.nasa.gov/) global meteorology, surface solar energy and climatology data.

For example, we can use `get_power()`, to obtain global rainfall (specifying `pars = "PRECTOT"`) or humidity (specifying `pars = "RH2M"`). 


```{r, eval = FALSE}
# library(nasapower)
# library(terra)
# library(viridis)
# 
# climate_avg <- get_power(community = "AG", pars = "PRECTOT",
#                          lonlat = "GLOBAL", temporal_average = "CLIMATOLOGY")
# 
# 
# library("rnaturalearth")
# map <- ne_countries(returnclass = "sf")
# 
# climate_box <- split(climate_avg, climate_avg$PARAMETER)
# climate_box <- lapply(climate_box, function(x){
#   x["PARAMETER"] <- NULL
#   x})
# climate_box <- lapply(X = climate_box, FUN = as.matrix)
# 
# #retrieving precipitation data using the above made climate_box() function
# PRECTOT <- terra::rast(climate_box$PRECTOT[,c(1:2, 15)],
#      crs = "+proj=eqc +lat_ts=0 +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs", type = "xyz")
# 
# #converting above raster object into a data.frame for mapping 
# PRECTOT_df <- as.data.frame(PRECTOT, xy = TRUE, na.rm = TRUE)
# rownames(PRECTOT_df) <- c()
# 
# 
# #plotting the graph
# ggplot() + 
#   geom_raster(data = PRECTOT_df, aes(x = x, y = y, fill = ANN)) +
#   geom_sf(data = map, inherit.aes = FALSE, fill = NA) + 
#   scale_fill_viridis() +
#   labs(title = "Rainfall in inches", fill = "Annual Rainfall")

```


## Vegetation and Land cover

`MODIStsp` is an R package for downloading and preprocessing time series of raster data from [MODIS](https://modis.gsfc.nasa.gov/data/dataprod/) data products.


Here, we use the ‘Vegetation Indices 16-Day L3 Global 250’ product with the product IDs MOD13Q1(Terra Product ID), and MYD13Q1(Aqua Product ID), but will be represented by `M*D13Q1` - the second character is replaced by an asterix(`*`) to identify both Terra and Aqua.

We can also download the ‘Land Cover Type Yearly L3 Global 500m’ product with the product ID `MCD12Q1` (Combined Aqua and Terra Product ID)

```{r}
# install.packages("MODIStsp")
# library(MODIStsp)
# library(rgeoboundaries)
# library(sf)
# 
# # Downloading the country boundary
# map_boundary <- geoboundaries("Kenya")
# 
# # Defining filepath to save downloaded spatial file
# spatial_filepath <- "map.shp"
# 
# # Saving downloaded spatial file on to our computer
# #st_write(map_boundary, paste0(spatial_filepath))
# ```
# 
# ```{r, eval = FALSE}
# MODIStsp(gui             = FALSE,
#          out_folder      = "VegetationData",
#          out_folder_mod  = "VegetationData",
#          selprod         = "Vegetation_Indexes_16Days_1Km (M*D13A2)",
#          bandsel         = "NDVI", 
#          user            = "mstp_test" ,
#          password        = "MSTP_test_01",
#          start_date      = "2020.06.01", 
#          end_date        = "2020.06.01", 
#          verbose         = FALSE,
#          spatmeth        = "file",
#          spafile         = spatial_filepath,
#          out_format      = "GTiff")
```

[Paula: I've set this to eval=FALSE for now to avoid filenotfound error, can we make this runnable ?, Andy]
```{r, eval=FALSE}
# # Reading in the downloaded NDVI raster data
# NDVI_raster <- raster("VegetationData/map/VI_16Days_1Km_v6/NDVI/MYD13A2_NDVI_2020_153.tif")
# 
# # Transforming the data
# NDVI_raster <- projectRaster(NDVI_raster, crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
# 
# # Cropping the data
# NDVI_raster <- raster::mask(NDVI_raster, as_Spatial(map_boundary))
# 
# # Dividing values by 10000 to have NDVI values between -1 and 1
# gain(NDVI_raster) <- 0.0001
# 
# # Converting the raster object into a dataframe
# NDVI_df <- as.data.frame(NDVI_raster, xy = TRUE, na.rm = TRUE)
# rownames(NDVI_df) <- c()
# 
# # Visualising using ggplot2
# ggplot() + 
#   geom_raster(data = NDVI_df, aes(x = x, y = y, fill = MYD13A2_NDVI_2020_153)) +
#   geom_sf(data = map_boundary, inherit.aes = FALSE, fill = NA) +
#   scale_fill_viridis(name = "NDVI") +
#   labs(title = "NDVI (Normalized Difference Vegetation Index)",
#        subtitle = "01-06-2020", x = "Longitude", y = "Latitude")

```

## Demographic and Health Survey (DHS)

The `rdhs` packages gives the users the ability to access and make analysis on the Demographic and Health Survey (DHS) data.
For example we can find out the trends in antimalarial use in Africa as follows.

```{r}
# install.packages("rdhs")
# library(rdhs)
# 
# # Make an api request
# resp <- dhs_data(indicatorIds = "ML_FEVT_C_AML", surveyYearStart = 2010, breakdown = "subnational")
# 
# # filter it to 12 countries for space
# countries  <- c("Angola","Ghana","Kenya","Liberia",
#                 "Madagascar","Mali","Malawi","Nigeria",
#                 "Rwanda","Sierra Leone","Senegal","Tanzania")
# 
# 
# library(ggplot2)
# ggplot(resp[resp$CountryName %in% countries, ],
#        aes(x = SurveyYear, y = Value, colour = CountryName)) +
#   geom_point() + geom_smooth(method = "glm") + 
#   theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
#   ylab(resp$Indicator[1]) + facet_wrap(~ CountryName, ncol = 6) 
```


## Malaria

The `malariaAtlas` package can be used to download, visualise and manipulate global malaria data hosted by the Malaria Atlas Project.

The package enables users to download the following types of data:

- parasite rate (PR) survey data (Plasmodium falciparum and Plasmodium vivax)
- vector occurrence data
- administrative boundary shapefiles to visualise data
- rasters covering a range of modelled outputs related to malaria research such as predicted malaria parasite prevalence

For example, we can download parasite rate survey data corresponding to Zimbabwe as follows.

```{r}
# install.packages("malariaAtlas")
# library(malariaAtlas)
# d  <- getPR(country = "Zimbabwe", species = "BOTH") 
# autoplot(d)
```