# Case study 2 {#case-study2}
Mapping real-time COVID cases

Hello, how are you doing.

# Template chapter {#template}

## Overall goal of the chapter
In this section we will describe what the overall aim of this chapter and target audience including:

- What they are already expected to know (their previous knowledge)
- What are the goals of the audience

<style>
div.green { background-color:#e3ffd9; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">
**Learning objectives** 

- What you will learn
- What you will be able to do at the end of this chapter
- What you will not learn & what to read instead
- Where to go from here
</div>

## Section 1 (topic 1) {#episode_1_template}
Each section developed within the chapter should contain a narrative to give a bit of a context and again demonstrate the target audience (if needed). Beyond it should provide topic-specific learning objectives and examples, which are to be intertwined with so-called *self-check exercises* which will allow the reader to assess his comprehension of the topic so far. 

<style>
div.purple { background-color:#e9d0f7; border-radius: 5px; padding: 20px;}
</style>
<div class = "purple">
**Exercise 1**: The exercise description to be in the coloured box & the small, control exercises should be placed at the end of the subsection.
</div>    


## Section 2 (topic 2) {#episode_2_template}
<style>
div.yellow { background-color:#fff5e6; border-radius: 5px; padding: 20px;}
</style>
<div class = "yellow">
**Shortcut**: Shortcuts to be in a yellow box.
</div>    

## Section 3 (topic 3) {#episode_3_template}


## Further resources
In this section we provide links to additional websites and sources that are useful in a given topic or package in the following way:

- [R website](r-project.org)
- [Documentation pages](rdocumentation.org)

## Summary
This closing section should show the reader what has been done in this chapter and how it links with the topics provided in the next chapter by a brief introduction to next topic.

## Exercise solutions

- Exercise 1
```{r, results='hide', fig.show='hide'}
#code to exercise 1 only, results hidden 
```

- Exercise 2
```{r, results='hide', fig.show='hide'}
#code to exercise 2 only, results hidden
```

## Putting into practice
[we are not decided if these sections necessary, may be too cluttered]

This final section is aimed at consolidating the knowledge and skills learned throughout the whole chapter, where the reader can put them into practice in a more complex working example with a new context. 