--- 
title: "The afrimapr book"
author: "The afrimapr team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "An introduction to making maps from African data in R."
---

# Welcome {-}
This is the in-progress book introducing map making in R using African data.

It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project running in 2020-21.

We welcome contributions, and will make avenues for contribution clearer soon.

In the meantime, do [get in touch](https://afrimapr.github.io/afrimapr.website/get-involved/).


# Preface {-}

In-progress chapters

1. Intro : aims of this book \@ref(intro)
1. Step through a simple example of process of mapmaking \@ref(steps1)
1. R recap - help ensure readers know foundations \@ref(rrecap)
1. Spatial data types focusing on real examples : cities, roads, admin boundaries, surfaces
1. Visualising spatial data part 1 : sf, mapview, tmap
1. Shiny interactive map applications
1. Sources of data, what is open data (could combine with chap 4)
1. Join non-spatial data onto spatial
1. Raster manipulation, e.g. summarising data by areas
1. OpenStreetMap data, what they are and how to use
1. Creating and editing spatial data
1. Visualising spatial data part 2, more advanced techniques
1. Case study 1 : health facility locations
1. Case study 2 ...


The version of the book you are currently reading was last updated at:

```{r}
Sys.time()
```

![Continuous integration (built on GitHub)](https://github.com/afrimapr/book/workflows/tic/badge.svg)

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
