--- 
title: "The afrimapr book"
author: "The afrimapr team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This book aims to provide an introduction to mapping with open data, with a focus on open access administrative and health data in Africa."
---

# Preface

Planned chapters

1. Intro : aims of this book \@ref(intro)
1. Step through a simple example of process of mapmaking [cake first, show achievable]
1. R recap - help ensure readers know foundations, data structures and manipulation
1. Spatial data types focusing on real examples : cities, roads, admin boundaries, surfaces
1. Visualising spatial data part 1 : sf, mapview, tmap
1. Shiny interactive map applications
1. Sources of data, what is open data (could combine with chap 4)
1. Join non-spatial data onto spatial
1. Raster manipulation, e.g. summarising data by areas
1. OpenStreetMap data, what they are and how to use
1. Creating and editing spatial data
1. Visualising spatial data part 2, more advanced techniques
1. Case study 1 : health facility locations
1. Case study 2 ...




- What the project is

- Link to main afrimapr website

- Who the book is for...

The version of the book you are currently reading was last updated at:

```{r}
Sys.time()
```

![Continuous integration (built on GitHub)](https://github.com/afrimapr/book/workflows/tic/badge.svg)

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
