--- 
title: "The afrimapr book"
author: "The afrimapr team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "An introduction to making maps from African data in R."
---

# Welcome {-}
This is the in-progress book introducing map making in R using African data.

It is part of the [afrimapr](https://afrimapr.github.io/afrimapr.website/) project running in 2020-21.

We welcome contributions, and will make avenues for contribution clearer soon.

In the meantime, do [get in touch](https://afrimapr.github.io/afrimapr.website/get-involved/).


# Preface {-}

[defined in index.Rmd]
[andy:links to chapters sometimes don't work for unknown reasons]
[https://bookdown.org/yihui/bookdown/cross-references.html]

In-progress chapters

1. Intro : aims of this book \@ref(intro)
1. Step through a simple example of process of mapmaking \@ref(steps1)
1. R recap - help ensure readers know foundations \@ref(rrecap)
1. Spatial data types focusing on examples \@ref(data-types)
1. Visualising spatial data part 1 : sf, mapview, tmap \@ref(visualising1)
1. Shiny interactive map applications \@ref(shiny)
1. Sources of data, what is open data \@ref(data-sources) 
1. Join non-spatial data onto spatial \@ref(joining)
1. Raster manipulation, e.g. summarising data by areas \@ref(raster)
1. OpenStreetMap data, what they are and how to use \@ref(openstreetmap)
1. Creating and editing spatial data \@ref(data-editing)
1. Visualising spatial data part 2, more advanced techniques \@ref(visualising2)
1. Case study 1 : health facility locations \@ref(case-study1)
1. Case study 2 \@ref(case-study2)


The version of the book you are currently reading was last updated at:

```{r}
Sys.time()
```

![Continuous integration (built on GitHub)](https://github.com/afrimapr/book/workflows/tic/badge.svg)

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
